---
title: "Intro to Data Visualization <br> in R for Research"
subtitle: "Part 1: A Data Driven Hope"
author: "Thomas Wise"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```



## Session Contents

- Intro to the Tidyverse
- Plotting in R, *plot* vs *ggplot2*
- Basics of ggplot2
  + Core Functions
  + Layering & Aesthetic Mapping 
  + Chart types (using Geoms)
  + Colours, Shapes and Size 
  + Other Aesthetic Details 
  
## Objectives and Transferable Skills 

<div style="float: left; width: 50%;">
- To be able to: 
  + Begin to generate simple visualizations from novel data 
  + Begin to utilize good coding and data visualization practices 
  + Begin to think constructively about data visualization 
</div>

<div style="float: right; width: 50%;">
- Core Skills: 
  + Data visualization in R
  + Exploratory Data Analysis using visualization 
  + Creative Problem solving and code debugging
</div>

  
## Session Layout

- 90 minutes long.
- Heavily focused on Practical, with some theory and visual demonstrations 
- When working on the practical content, feel free to work on your own, in pairs, in groups! 

## Schedule

<div style="float: left; width: 50%;">
**Introduction** (15 minutes)

- Intro to the Tidyverse  
- *plot* vs *ggplot*  

**Part 1** (30 minutes)

- Basics of ggplot2 (1)  
  + Core Functions & Structure  
  + Layering and Aesthetics 
  + Legends & Labels
  
**Break** (10 minutes)  
</div>

<div style="float: right; width: 50%;">
**Part 2** (25 minutes)  

- Basics of ggplot2 (2)  
  + Chart Types (Geoms)
  + Shapes, Colours and Size  
  + Other Aesthetic Details  
  
**Review & Conclusion** (10 minutes)  

- Review our Session  
- Discuss content of next session  
- Sharing examples  
</div>

# Introduction 

## The Tidyverse 

- A collection of R Packages designed specifically for data science.
- This includes packages which are considered *core* in everyday data analysis.
- These are designed to encourage uniform and consistent practice within data science programming. 

- Packages include  
  + *ggplot2*: Used for generating graphics 
  + *dplyr*: Used for data manipulation 
  + *tidyr*: Used for data cleaning and tidying 
  + *readr*: Used to read and load data (like .csv, .tsv)
  + And many, many more. 
  
## R for Data Science 

- The *bible* for learning the tidyverse is [R for data science](https://r4ds.had.co.nz/)
- This covers a general introduction to Tidyverse packages and goes from complete beginners through to basic skills. 
- One useful tool, are the [RStudio Cheatsheets](https://rstudio.com/resources/cheatsheets/)

## Ways to plot in R, plot vs ggplot2

<div style="float: left; width: 45%;">

```{r, fig.width = 4}
plot(x = diamonds$carat,
     y = diamonds$price)

```

</div>

<div style="float: right; width: 45%;">

```{r, fig.width = 4, fig.height = 4}
ggplot(data = diamonds, 
       mapping = aes(x = carat, 
                     y = price)) + 
  geom_point()

```

</div>

## Plot vs ggplot2

<div style="float: left; width: 45%;">
**Plot**

- Simple and Easy Method,
- Can be used quickly without much coding experience,
- Not useful for visualizing complex data or generating nice plots! 


</div>
<div style="float: right; width: 45%;">

**ggplot2**

- Complex, but highly adaptive, 
- Layer based, and can be used for lots of different graph types,
- Extremely useful for visualizing complex data and generating nice pleasing plots,

</div>

# Part 1

## Basics of ggplot2

- Your basic ggplot2 function contains the following: 
- Core Function

```{r, eval = FALSE}
ggplot()
```

- Plotting Specification (geom)

```{r, eval = FALSE}
geom_point()
geom_histogram()
geom_line()
```

- Mapping Function & Data Specification

```{r, eval = FALSE}
mapping = aes(x, y)
data = data
```

## ggplot2 in action 

We can see this in action through the code: 

```{r}

ggplot(data = diamonds, 
       mapping = aes(y = carat, x = price)) + 
  geom_point()

```

## Our Data 

- Kaggle data set [Online Video Gaming and Anxiety](https://www.kaggle.com/divyansh22/online-gaming-anxiety-data). 
- 15,000 observations across 55 variables, for this workshop we will use a sample of 1,000
- Selected for its Clinical Variables and High Usability. 

- In general, we will explore the factors which influence an individuals anxiety score (GAD7 and SPIN), including the age of the player, platform, time spent playing as well as other factors. 

## Exercise 1: Starting basic 

Lets introduce this data set through covering the basics of ggplot2. 
Lets firstly produce a histogram examining the distribution of the amount of hours played. 

Data can be imported using the code provided in the worksheet! 
You can either use the template provided in the worksheet or have a go without it! 

## Exercise 1: Solution

```{r, warning = FALSE, message = FALSE}
## Import the data
gamingstudy_data <- read_csv("data/gamingdata_samp.csv")

## Plot the data 
ggplot(data = gamingstudy_data, 
       mapping = aes(x = Hours)) + 
  geom_histogram()

```

## Exercise 2: Aesthetic Mapping 

Next we can consider plotting a histogram of GAD scores (*GAD_T*), we can see that we need to consider aspects such as binwidth. 

```{r}
ggplot(data = gamingstudy_data, 
       mapping = aes(x = GAD_T)) + 
  geom_histogram(bins = 30)

```

Right now, the histogram uses the parameter *bins = 30*, however we can specify the *binwidth* to improve the plot. 

## Exercise 2: Solution 

```{r}

## Plot the data 
ggplot(data = gamingstudy_data, 
       mapping = aes(x = GAD_T)) + 
  geom_histogram(binwidth = 1)

```

## Excerise 3: Additional Aesthetic Mapping & Layers



## Exercise 3: Solution

```{r}
## Plot the data 
colours <- c("GAD" = "blue", "SPIN" = "red")

ggplot(data = gamingstudy_data) + 
  geom_histogram(mapping = aes(x = GAD_T, colour = "GAD"), binwidth = 1, alpha = 0.1) + 
  geom_histogram(mapping = aes(x = SPIN_T, colour = "SPIN"), binwidth = 1, alpha = 0.1) + 
  labs(colour = "Legend") + 
  scale_color_manual(values = colours)

```

## Exercise 4: Labels & Legends


## Exercise 4: Solution 

```{r}
## Plot the data 
colours <- c("GAD" = "blue", "SPIN" = "red")

ggplot(data = gamingstudy_data) + 
  geom_histogram(mapping = aes(x = GAD_T, colour = "GAD"), binwidth = 1, alpha = 0.1) + 
  geom_histogram(mapping = aes(x = SPIN_T, colour = "SPIN"), binwidth = 1, alpha = 0.1) + 
  labs(x = "Anxiety Measure",
       y = "Frequency", 
       colour = "Legend") + 
  scale_color_manual(values = colours)

```

## Exercise 5: Chart Types & Aesthetics 


## Exercise 5: Soultion 

```{r}



```


# Break (10 minutes)

# Part 2

## Chart Types 

## Exercise 6: Different Types of Charts 

## Exercise 6: Solution 

## Exercise 7: Colours

## Exercise 7: Solution

## Exercise 8: Shapes and Sizes 

## Exercise 8: Solution 

## Exercise 9: Additional Aesthetic Layers & Details (2)

## Exercise 9: Solution 

# Review and Conclusion 

## Review 

## Plan for Next Session






